Short-term neuronal and synaptic plasticity act in synergy for deviance detection in spiking networks

### 2022-12-28 paper-1/manuscript.tex
Rev sec Results:F8

We had good reasons for this assumption: Firstly, notice that the ablated model with STD only produced only a positive SSA index (i.e., the target response in msc was greater than in std), but not a significantly positive DDI (i.e., no difference between target dev and msc responses). Secondly, we note that updates to the spike-triggered depression variable $x_j$ (cf. Equation \ref{eqn-xsyn}) would decay to a negligible $e^\frac{-1 s}{\tau_x} = 0.0013$ times the original value within two trials, making it unlikely that inputs separated by intervening trials could be "remembered" in the network. Conversely, the frequently presented stimulus B could lead to a build-up of depression strong enough to suppress its own response. Referring back to Figure \ref{fig3}b, we see that there was indeed a significant, if small, effect of presentation frequency even in the ablated model with STD only, yielding a positive SSA index on average.
% To show this in the sample network, we correlated the number of spikes in dev target trials with the spike-triggered depression variable $x_j$ at the start of later trials (cf. Equation \ref{eqn-xsyn}) across excitatory neurons. We found that, while the correlation was very clear in the trial immediately following target (Pearson's $\rho$ = -0.984, p = 0; n = 800), it was almost completely lost by the next trial ($\rho$ = -0.0525, p = 0.138) under the influence of both decay and unrelated activity in the intervening trial.

To show that STD being carried over between trials was responsible for the reduction of non-target dev pulses relative to msc, we needed to estimate the extent to which activity of a given neuron (or, alternatively, its membrane potential $V_m$) was affected by changes in STD. Unlike $V_{TA}$, which directly affected the neuron to which the value belongs, $x_j$ exerted its effects on the other side of a synapse. Therefore, we devised a measure $D$ to estimate the postsynaptic impact of STD in terms of membrane potential lost, mirroring the effect of $V_{TA}$ in the sense that decreasing $V_m$ and increasing the firing threshold both increase the depolarization necessary to bring a neuron to fire.
A measure of STD is necessarily activity-dependent -- a postsynaptic neuron's membrane potential is driven only by presynaptic partners that fire -- so it must be calculated based on observed activity. Here, we used the average trial response $R_j^S$ of neuron $j$ to a given stimulus $S$, in number of spikes, as the presynaptic drive. Before depression, a single excitatory presynaptic spike evokes a postsynaptic potential (EPSP) peaking at approximately $k = 1.4$ mV at rest (recall that $w = 1$ for all synapses), and STD linearly scales this EPSP by presynaptic $x_j$. Thus, we estimated the average peak EPSP evoked by a spike in presynaptic neuron $j$ as $R_j^S k x_j^S$. For this estimate, we ignored added depression due to repeated spikes within a trial, instead defining $x_j^S$ as the average depression at the start of trials. Finally, summing over presynaptic neurons, we estimated the peak EPSP lost to depression in a given neuron $i$ as
\begin{equation}
    D_i^S = \sum_{j \in pre} R_j^S k (1-x_j^S)
\end{equation}
Notice that $D$ is measured in volts due to scaling with $k$, and is therefore roughly comparable to $V_{TA}$, albeit more approximate. Lastly, since we contrasted this measure between conditions (see below), we aimed to avoid effects purely due to differences in presynaptic activity. Therefore, we calculated $R^S$ as the average response across relevant trials in all conditions, leaving the contrast entirely in the hands of $x^S$.

\begin{figure*}%
    \centering
    \includegraphics{fig8.pdf}
    \caption{}
    \label{fig8}
\end{figure*}
\textbf{Fig. 8} In the full model, STD reduces dev non-target and increases dev target responses relative to msc.
\textbf{a} Average depression $D$ in target (A) trials (top row) and B trials (bottom row) in the dev (left) and msc sequence (center), and their contrast (right). The color scale is shared across columns (and between dev and msc) for direct comparability.
\textbf{b} Statistics across networks and stimuli of the contrast (dev - msc) in $D$ in A trials (left) and B trials (right)
\textbf{c} Relationship across neurons of the sample network between the contrast (dev - msc) in $D^S$ and the contrast in average number of spikes fired in $S$ trials, and associated linear regression. Left, target trials; right, B trials.
\textbf{d} Network-wide mean contrast (dev - msc) in $D^S$ across networks and stimuli, plotted against the Pearson correlation coefficients $\rho$ of the relationship shown in panel c ($\Delta D^S$ vs response size contrast across neurons). Each marker corresponds to one network and target stimulus. Red markers indicate a significant negative correlation (one-sided t-test, significance level 0.05). Inset histograms show the distribution of the corresponding values, with orange portions indicating the 95\% confidence interval of the median, calculated by bootstrap with 10000 iterations.

As noted above, we hypothesized that, in the full model, STD is directly responsible only for a reduction of non-target dev trials, relative to msc, whereas differences between target dev and msc trials are not a result of STD directly. To show this, we calculated $D$ for A and B trials, and contrasted it across dev and msc conditions as with $V_{TA}$ before.
In Figure \ref{fig8}a, we see that target trials (top) experienced depression across large parts of the sample network in both conditions. The msc sequence, however, clearly evoked greater depression, as borne out in the contrast (dev - msc), which was negative throughout. Conversely, depression in B trials (bottom) was spatially more clustered around the stimulation site in both conditions. In the msc sequence, we can attribute the difference between A and B primarily to the response size: The context was the same, but responses to B tended not to spread out as much in this network. However, in the dev sequence, the B response was limited almost exclusively to a very local patch around the stimulation site (see also Figure \ref{fig6}a), with depression therefore similarly limited. As a result, the contrast (dev - msc) showed slightly reduced $D$ in the periphery, but strikingly elevated $D$ near the stimulation site, indicating greatly increased synaptic depression in dev.

To show how robust these effects were, we turn to statistics across networks and stimuli. We calculated the trial-average depression values $D^A$ and $D^B$ in dev and msc conditions, subtracted them, then calculated the mean over neurons. The resulting contrasts are summarized in Figure \ref{fig8}b. We note first that neurons were subject to more depression in B trials than in A trials in the dev sequence (-4.47, p = 9.02e-06; one-sided t-test, n = 120, data not shown), but not in the msc sequence (by definition). Furthermore, $D^B$ was greater in dev than msc (8.02, p = 4.05e-13). Both findings confirm that STD carried over between the frequent standard trials, but not, or less so, between the less frequent control and deviant trials.
However, we also found that $D^A$ was greater in msc than dev (-7.54, p = 5.04e-12; one-sided t-test, n = 120), suggesting that STD did exert stimulus-independent suppression in the msc sequence, as is clearly seen in the sample network, and similar to what we described earlier for TA.

Finally, to show that $D$ reflects suppression of a neuron's activity, we quantified the relationship between the contrasts in $D$ and the corresponding response size contrast. In the sample network (Figure \ref{fig8}c), we see that neurons with a greater decrease in $D$ from msc to dev had correspondingly greater response increase, and vice versa. Across both A and B trials, this relationship was captured well by a linear regression (A: Pearson's $\rho$ = -0.541, p = 2.28e-77, B: $\rho$ = -0.186, p = 1.61e-09; one-sided t-test, n = 1000). These findings were borne out across most networks and stimuli (Figure \ref{fig8}d), showing that $D$ captures the impact of STD not only on $V_m$, but also on the resulting output spikes.

To sum up, in the full model, STD acted in much the same way as TA, causing both a reduction of the response to the frequently presented standard stimuli (dev B) by adaptation, and an increase in the response to deviant stimuli (dev A) by relief from adaptation, relative to control. Yet, as we showed in Figure \ref{fig7} and in the previous section, a decrease in dev B responses also entailed lower $V_{TA}$, and higher dev A response by consequence. Thus, adding STD to the model increased the DDI both directly, as outlined at the start of this paragraph, and indirectly, by altering the pattern of TA by way of changed responses.

>>>>

We had good reasons for this assumption: Firstly, referring back to Figure \ref{fig3}, notice that the ablated model with STD only produced only a positive SSA index (i.e., the target response in msc was greater than in std), but not a significant DDI (i.e., no difference between target dev and msc responses). Secondly, we note that updates to the spike-triggered depression variable $x_j$ (cf. Equation \ref{eqn-xsyn}) would decay to a negligible $e^\frac{-1 s}{\tau_x} = 0.0013$ times the original value within two trials, making it unlikely that inputs separated by intervening trials could be "remembered" in the network. Conversely, the frequently presented stimulus B could lead to a build-up of depression strong enough to suppress its own response.
To show that the network does not carry information beyond the most recent trial in the sample network, we correlated the number of spikes in dev target trials with the spike-triggered depression variable $x_j$ at the start of later trials (cf. Equation \ref{eqn-xsyn}) across excitatory neurons. We found that, while the correlation was very clear in the trial immediately following target (Pearson's $\rho$ = -0.984, p = 0; n = 800), it was almost completely lost by the next trial ($\rho$ = -0.0525, p = 0.138) under the combined influence of decay and unrelated activity in the intervening trial.

To show that STD being carried over between repeated trials was responsible for the reduction of non-target dev pulses relative to msc, however, we needed to estimate the extent to which activity of a given neuron (or, alternatively, its membrane potential $V_m$) was affected by changes in STD. Unlike $V_{TA}$, which directly affects the neuron to which the value belongs, $x_j$ exerts its effects on the other side of a synapse. Therefore, we devised a measure $D$ to estimate the postsynaptic impact of STD in terms of membrane potential lost, mirroring the effect of $V_{TA}$ in the sense that decreasing $V_m$ and increasing the firing threshold both increase the depolarization necessary to make a neuron fire.
A measure of STD is necessarily activity-dependent -- a postsynaptic neuron's membrane potential is driven only by presynaptic partners that fire -- so it must be calculated based on observed activity. Here, we used the average trial response $R_j^S$ of neuron $j$ to a given stimulus $S$, in number of spikes, as the presynaptic drive. Before depression, a single excitatory presynaptic spike evokes a postsynaptic potential (EPSP) peaking at approximately $k = 1.4$ mV at rest (recall that $w = 1$ for all synapses), and STD linearly scales this EPSP by presynaptic $x_j$. Thus, we estimated the average peak EPSP evoked by a spike in presynaptic neuron $j$ as $R_j^S k x_j^S$. For this estimate, we ignored added depression due to repeated spikes within a trial, instead defining $x_j^S$ as the average depression at the start of trials. Finally, summing over presynaptic neurons, we estimated the peak EPSP lost to depression in a given neuron $i$ as
\begin{equation}
    D_i^S = \sum_{j \in pre} R_j^S k (1-x_j^S)
\end{equation}
Notice that $D$ is measured in volts due to scaling with $k$, and is therefore roughly comparable to $V_{TA}$, albeit more approximate. Lastly, since we contrasted this measure between conditions (see below), we aimed to avoid effects purely due to differences in presynaptic activity. Therefore, we calculated $R^S$ as the average response across $S$ trials in all conditions, leaving the contrast entirely in the hands of $x^S$.

\begin{figure*}%
    \centering
    \includegraphics{fig8.pdf}
    \caption{}
    \label{fig8}
\end{figure*}
\textbf{Fig. 8} In the full model, STD reduces dev non-target and increases dev target responses relative to msc.
\textbf{a} Average depression $D$ in target (A) trials (top row) and B trials (bottom row) in the dev (left) and msc sequence (center), and their contrast (right) in the sample network. The color scale is shared across columns (and between dev and msc) for direct comparability.
\textbf{b} Relationship across neurons of the sample network between the contrast (dev - msc) in $D^A$ and the contrast in the target response, and associated linear regression.
\textbf{c} Median contrast (dev - msc) in $D^B$ in the first 100 neurons to respond to B, and in the remaining, late portion of the network, across networks and stimuli.
\textbf{d} Relationship across networks and stimuli between the median contrast in $D^B$ among the early portion and the median contrast in the response to B among the entire network, and associated linear regression.
\textbf{e} Network-wide median contrast in $D^A$, plotted against its Pearson correlation coefficient with the corresponding contrast in the target response, across networks and stimuli. The plot shows both that the negative correlation shown in panel b is robust (most networks are plotted with red markers, i.e. significantly negatively correlated), and that $\Delta D^A$ is less than 0 on average (see the inset histogram on the horizontal axis)

As noted above, we hypothesized that, in the full model, STD is directly responsible only for a reduction of non-target dev trials, relative to msc, whereas differences between target dev and msc trials are not a result of STD directly. To show this, we calculated $D$ for A and B trials, and contrasted it across dev and msc conditions as with $V_{TA}$ before.
In Figure \ref{fig8}a, we see that target trials (top) experienced depression across large parts of the sample network in both conditions. The msc sequence, however, clearly evoked greater depression, as borne out in the contrast (dev - msc), which was negative throughout. Conversely, depression in B trials (bottom) was spatially more clustered around the stimulation site in both conditions. In the msc sequence, we can attribute the difference between A and B primarily to the response pattern: The context was the same, but responses to B tended not to spread out as much in this network. However, in the dev sequence, the B response was limited almost exclusively to a very local patch around the stimulation site, with depression therefore similarly limited. As a result, the contrast (dev - msc) showed slightly reduced $D$ in the periphery, but strikingly elevated $D$ near the stimulation site, indicating greatly increased synaptic depression in dev.

To show that $D$ captures not only changes in $V_m$, by definition, but also in the resulting response, we correlated the contrast (dev - msc) in $D^A$ against the target response contrast across neurons of the sample network. Given the relatively large responses in the contrasted conditions, and the infrequent presentation of these trials, allowing causality to mostly flow from $D$ to response, rather than also in reverse, we are confident that this relationship roughly reflects a causal influence. As shown in Figure \ref{fig8}b, the relationship was clearly negative (Pearson's $\rho$ = -0.541, p = 2.28e-77; Wald test, n = 1000), indicating that $D$ reduces responses as expected.

Having gained a rough intuition of how individual neurons are affected by STD, we now turn to its effect on the network, working with statistics across networks and stimuli. First, in Figure \ref{fig8}c, we recapitulate the logic established in Figure \ref{fig6}: Given the spatial clustering of $D^B$, we surmised a similar local suppression effect as with TA. We found that, indeed, neurons responding early to stimulation in B showed greater $D^B$ in dev than msc (T = 7.26e+03, p = 1.01e-21; one-sided Wilcoxon, n = 120), while the late portion had slightly reduced $D^B$ (T = 2.24e+03, p = 0.000133). We can relate this directly to the non-target response contrasts in the early and late portions shown in Figure \ref{fig7}a, which followed a similar pattern, and take this to indicate that the increased presentation of B in dev causes local suppression in terms of $D$. % Possible concern: The relationship dXnotA ~ dD^B is not clearly established.

To show that increased local suppression by STD reduces the non-target response, we correlated the median contrast in $D^B$ among the early portion of networks with the median contrast of the total response to B. As we show in Figure \ref{fig8}d, higher $\Delta D^B$ was clearly predictive of a lower network response ($\rho$ = -0.581). This confirms again the integrity of the measure $D$, but also that STD-mediated local suppression played a role in reducing the non-target response in the dev sequence. % Possible concern: The equivalent hasn't been shown in TA (Fig6), but likely could be (in the same fashion; dTA ~ dXB)

Finally, given the very clear contrast in $D^A$ in the sample network, we considered the possibility that, in the context of the full model, STD could also evoke global suppression in the msc sequence, contrary to our stated hypothesis. If so, we should expect to see less $D^A$ in dev than msc. To confirm this, we analysed the contrast (dev - msc) in $D^A$ across networks and stimuli, and found its network median to be robustly negative (T = 955, p = 1.23e-12; one-sided Wilcoxon, n = 120). In addition, as shown in Figure \ref{fig8}e, we correlated the contrast in $D^A$ with the target response contrast in each network, and found that weaker depression in dev predicted a greater response in most networks (T = 724, p = 1.37e-14), though the median correlation coefficient $\rho$ was only -0.282. We note that calculating this same correlation in the model with STD only yields no significantly negative dependence (T = 3.99e+03, p = 0.828; see Figure \ref{sfig8}). Taken together, we cautiously conclude that, while there is clear evidence for greater global suppression in msc than dev in terms of $D$, this difference appears to manifest weakly in activity differences.

To sum up, in the full model, STD acted in much the same way as TA, causing both a reduction of the response to the frequently presented standard stimuli (dev B) by local suppression, and an increase in the response to deviant stimuli (dev A) by relief from global suppression, relative to control. Yet, as we showed in Figure \ref{fig7} and in the previous section, a decrease in dev B responses also entailed lower $V_{TA}$, and higher dev A responses by consequence. Thus, adding STD to the model increased the DDI both directly, by reducing global suppression, and indirectly, by altering the pattern of TA by way of local suppression of B.

We note that the direct and indirect effect of STD on global suppression appear to be on the same order of magnitude. To see this, compare the change in the $V_{TA}$ contrast at the start of A trials as a result of adding STD to the model (Figure \ref{fig7}d, bottom) with the corresponding $\Delta D^A$ (Figure \ref{fig8}e), both of which are on the order of $\pm$ 0.1 mV. Conversely, notice that, since TA and STD work in similar fashion to increase the dev target response, it is very plausible that TA likewise increases the effect of STD, by enhancing the STD-mediated local suppression of B and thereby increasing the contrast in STD-mediated global suppression.
